\begin{technical}
{\Large\textbf{Coordinate Reversal and Causal Geometry}}\\[0.7em]

\noindent\textbf{Schwarzschild Metric and the Horizon}\\[0.5em]
The Schwarzschild metric describes spacetime outside a static mass \(M\). In spherical coordinates \((t, r, \theta, \phi)\), it takes the form:
\begin{align}
\mathrm{d}s^2 =\ 
& -\left(1 - \frac{2GM}{rc^2}\right)c^2\,\mathrm{d}t^2 \nonumber \\
& + \left(1 - \frac{2GM}{rc^2}\right)^{-1}\mathrm{d}r^2 \nonumber \\
& + r^2\,\mathrm{d}\theta^2 + r^2\sin^2\theta\,\mathrm{d}\phi^2.
\end{align}

The Schwarzschild radius is defined as:
\[
r_s = \frac{2GM}{c^2}.
\]
At \(r = r_s\), \(g_{tt} = 0\), \(g_{rr} \to \infty\). This signals a breakdown in coordinates, not in geometry.

\vspace{0.7em}
\noindent\textbf{Time Dilation and Gravitational Redshift}\\[0.5em]
A static observer at fixed radius \(r > r_s\) experiences proper time:
\begin{equation}
\mathrm{d}\tau = \sqrt{1 - \frac{2GM}{rc^2}}\,\mathrm{d}t.
\end{equation}
As \(r \to r_s\), \(\mathrm{d}\tau/\mathrm{d}t \to 0\). Distant clocks appear to tick normally, but local clocks slow near the horizon.

Photons emitted at \(r_{\text{em}}\) and received at \(r_{\text{obs}} \to \infty\) undergo redshift:
\begin{equation}
1 + z = \left(1 - \frac{2GM}{r_{\text{em}}c^2}\right)^{-1/2}.
\end{equation}
As \(r_{\text{em}} \to r_s\), \(z \to \infty\). Light from near the horizon becomes infinitely stretched and fades from view.

\vspace{0.7em}
\noindent\textbf{Finite Infall and Apparent Freezing}\\[0.5em]
A freely falling observer from rest at infinity reaches radius \(r\) in proper time:
\begin{equation}
\tau(r) = \frac{2}{3}\frac{r^{3/2}}{\sqrt{2GM}}.
\end{equation}
The total proper time to reach the horizon is finite. The infaller feels no discontinuity; the horizon is not a physical surface.

However, to a distant observer, the infaller appears to freeze at \(r = r_s\), due to the divergence of coordinate time:
\[
t(r) \to \infty \quad \text{as} \quad r \to r_s.
\]

\vspace{0.7em}
\noindent\textbf{Coordinate Inversion Below the Horizon}\\[0.5em]
Inside the horizon (\(r < r_s\)), the signs of metric components reverse:
\[
g_{tt} > 0, \qquad g_{rr} < 0.
\]
This means \(r\) is now a timelike coordinate, and \(t\) becomes spacelike.

The physical implication is that movement in \(r\) becomes mandatory. All future-directed timelike paths lead to smaller \(r\), ending at the singularity \(r = 0\). The singularity is not “a place inside” but a moment in the infaller’s proper future.

This inversion is not a coordinate trick — it reflects the actual causal structure of the geometry. Any attempt to "hover" or remain at constant \(r\) is no longer physically possible once inside the horizon.

\vspace{0.7em}
\noindent\textbf{Light Cones and Irreversibility}\\[0.5em]
At the horizon, outgoing light rays remain on the surface:
\[
\frac{\mathrm{d}r}{\mathrm{d}t} = 0 \quad \text{for outgoing null rays at } r = r_s.
\]
Below the horizon, all light cones tip inward. Future lightlike and timelike paths are directed toward decreasing \(r\). There is no direction within the cone that leads to increasing radius.

This defines the event horizon as a one-way temporal boundary: a surface from which causal influence cannot escape outward.

\vspace{0.7em}
\noindent\textbf{Summary: Time as Radius}\\[0.5em]
Outside the horizon, \(r\) measures spatial distance, and \(t\) is global time. Inside, the reverse is true: \(r\) becomes a clock, and all worldlines evolve toward smaller \(r\) as they progress into the future.

This behavior illustrates why the singularity is not a location, but an inevitable moment. The black hole interior is not a region to explore, but a domain of future-directed inevitability in curved spacetime.

\vspace{0.7em}
\noindent\textbf{References}\\
Misner, C. W., Thorne, K. S., Wheeler, J. A. (1973). \textit{Gravitation}.\\
Carroll, S. M. (2004). \textit{Spacetime and Geometry}.\\
Wald, R. M. (1984). \textit{General Relativity}.
\end{technical}
